Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-02 appendix.tex
Hopefully my last stupid questions.
terminology; going through comments and writing out steps explicitly

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
%EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
32 times likelihoods for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
An analogous statement holds for $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times likelihoods on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
Recall that
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \ge 0$.
The remaining variables follow similar arguments.

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{x}_2$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have

@@

@@

0 < b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{2}^U,

@@

0 < b_{1} \le \frac{x^*}{1-(x^*)^2} =: b_{1}^U,

@@

0 < b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{12}^U,

@@

\ell'_{L}(w) =  -\frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w} + \frac{1}{1+w}

@@

Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1]$, as shown in the previous proof, provided $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$.

@@

@@

%das: @EM did you say you wanted to get rid of these paragraphs? we never show there is a case where Fels > InvFels, just where it's equal...
%EM It doesn't seem necessary to me. If we shrink the internal branch to zero, the two topologies match, right? It just seems like we need a sentence or two expressing that.
As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
As we are interested in a lower bound, assume $\boldsymbol\xi=\boldsymbol\xi_0$, i.e., $\emptyset$ is the maximal ancestral state split for all site splits.
In this case, the conditional log-likelihood for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same conditional log-likelihood for both topologies.

Assume $w=1$.
In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the marginal log-likelihood is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
The unconstrained maximum likelihood value for $\tau_2$ is greater than or equal to the value of the likelihood given these two constraints.
We have
\begin{align*}
\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi) \ge \\
&\max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t : t^*\in\mathcal{T}_0} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi),
\end{align*}
and thus the incorrect topology achieves a likelihood value greater than or equal to the correct one.

>>>>

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $x$ is $(1-x)/2$, while the probability of a character remaining the same is $(1+x)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Conditional likelihood computations for all site splits and ancestral state splits are in Table~\ref{tab:farris_likelihoods} for the InvFels tree.

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
%EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
%das: went with "conditional likelihood"
32 times conditional likelihood values for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the conditional likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

Thus, we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$.
An analogous statement holds for $y_2 \ge y_1$, with these conditional likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times conditional likelihood values on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the conditional likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
Recall that
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
As we exclude the infinite branch length cases by letting $x_1,y_1,x_2,y_2,w > 0$, we have the strict inequality $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$.
The remaining variables follow similar arguments.

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal conditional likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given as the log of those in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{x}_2$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1 > 0$ and $x_2 \ge x_1 > 0$, we have

@@

%das: The lower bound for b_{\emptyset} is strict because we assume x_1,...,w > 0, but for the other three they aren't strict since x_1 could equal x_2 and/or y_1 could equal y_2. This doesn't affect any later results, though, b/c b -> 0 just makes that term drop out.

@@

0 \le b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{2}^U,

@@

0 \le b_{1} \le \frac{x^*}{1-(x^*)^2} =: b_{1}^U,

@@

0 \le b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} =: b_{12}^U,

@@

\ell'_{L}(w) :=  -\frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w} + \frac{1}{1+w}.

@@

%EM Where is this shown in the previous proof?
%das: wrote out more explicitly. It's similar to the argument starting at L612.
Each term in $\ell'_{L}(w)$ is of the form
f(w) = \frac{p}{1/b + w} = \frac{p\cdot b}{1 + b\cdot w}
with $0 \le p \le 1$.
Assume $b_{\emptyset}^U, b_{2}^U, b_{1}^U, b_{12}^U < 1$, so that $-1 < b < 1$.
Since $b > -1$ and $w\in(0,1]$, $1+b\cdot w > 0$ and, thus, each term in $\ell'_{L}(w)$ is a decreasing function in $w$.
%EM I agree that this condition shows that the lower bound is increasing on the entire interval. But how do we get to a conclusion about $\ell(.)$ itself?
%das: The derivative of \ell, i.e., \ell', is a decreasing function on [0,1]. The value of \ell' at 1 is positive, thus, \ell' > 0 on all of [0,1]. Since \ell' > 0 on all of [0,1], \ell is increasing on all of [0,1] (I think via MVT?).

@@

%EM Hm... it seems like we need to exclude having the b's very close to 1. How do we do that?
%das: I thought that was taken care of by the condition above: $b_{2}^U, b_{1}^U, b_{12}^U < 1$. It should also be taken care of by the below inequality, too, right? Otherwise as b -> 1 this will go to -\infty, so the below will fail. Do you think they need to be well-separated from 1 instead of just strictly less than 1?

@@

%das: @EM did you say you wanted to get rid of these paragraphs? we never show there is a case where Fels > InvFels, just where it's equal...
%EM It doesn't seem necessary to me. If we shrink the internal branch to zero, the two topologies match, right? It just seems like we need a sentence or two expressing that.
%EM Thoughts here?
%das: I kept the one paragraph showing they're equal, but maybe even that's too much?
As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
It is easy to see that assuming $\emptyset$ for each maximal ancestral state split results in the same conditional log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\boldsymbol\xi_0, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}
